\chapter{Athuls Contributions}

\section{Extended Literature Review}

\subsection{Early Detection and Temporal Dynamics}

A key limitation of content-based misinformation detection models is their reliance on fully propagated social context, which may not be available in early detection scenarios. Liu and Wu \cite{Liu2020} address this challenge through FNED, a deep network designed for fake news early detection on social media. FNED models the sequential behaviour of crowd responses over time, combining textual features with user-profile signals to classify news veracity within minutes of initial publication. Evaluated on Twitter and Weibo, FNED achieves over 90\% accuracy within the first five minutes of propagation, demonstrating that temporal behavioural patterns alone carry strong classification signals. This finding is particularly relevant to the MACE framework's Stage 2 experimental design, where early detection robustness under incomplete propagation data represents an important but underexplored evaluation dimension.

\subsection{Bi-Directional Propagation Modelling}

Expanding on graph-based approaches to social context modelling, Bian et al.\ \cite{Bian2020} propose Bi-Directional Graph Convolutional Networks (Bi-GCN) for rumour detection on social media. Unlike unidirectional propagation models, Bi-GCN simultaneously encodes both the top-down spreading pattern and the bottom-up aggregation of responses within a reply tree structure. Evaluated on Twitter15, Twitter16, and Weibo — the same benchmark datasets used in the MACE Stage 1 replication — Bi-GCN demonstrates that modelling bidirectional propagation significantly improves classification performance over prior graph-based baselines. This work provides additional motivation for the retention of the DANES and GETAE social context branches in the MACE framework, as these architectures similarly leverage structured propagation signals that single-direction models fail to capture fully.

\subsection{Image-Text Similarity and Cross-Modal Consistency}

Zhou et al.\ \cite{Zhou2020} introduce SAFE, a Similarity-Aware Multi-modal Fake News Detection framework that explicitly models the relationship between textual and visual content. SAFE extracts semantic representations from both text and images and computes a similarity score to identify inconsistencies between modalities — a common characteristic of manipulated or misleading posts. Evaluated on a Twitter-based dataset, SAFE outperforms prior unimodal and multimodal baselines, achieving strong accuracy while providing interpretable similarity scores. This approach is directly relevant to the multimodal content branch of MACE, as SAFE's cross-modal consistency modelling offers a complementary perspective to HMCAN's hierarchical attention fusion. Integrating similarity-based signals could further strengthen MACE's content branch, particularly for detecting posts where visual and textual content are semantically misaligned.

\subsection{Multimodal Variational Representation Learning}

Khattar et al.\ \cite{Khattar2019} propose MVAE, a Multimodal Variational Autoencoder for fake news detection that learns a shared latent representation from both textual and visual modalities through a bimodal variational autoencoder coupled with a binary classifier. Unlike discriminative models that rely on supervised feature extraction, MVAE employs generative modelling to capture the joint distribution of multimodal content, enabling the detection of subtle cross-modal inconsistencies that may not surface through attention-based fusion alone. Evaluated on Twitter and Weibo datasets, MVAE outperforms state-of-the-art methods by margins of approximately 6\% in accuracy and 5\% in F1 score. As a generative multimodal baseline, MVAE provides a useful point of comparison for the MACE framework's content branch evaluation, particularly in ablation settings where the contribution of visual features is isolated from social context signals.

\subsection{Out-of-Context Multimodal Misinformation}

Luo et al.\ \cite{Luo2021} present NewsCLIPpings, a large-scale benchmark dataset for out-of-context multimodal misinformation detection, constructed by automatically pairing real news images with semantically related but contextually mismatched captions. Unlike datasets where both text and image are fabricated, NewsCLIPpings specifically targets a prevalent real-world misinformation strategy in which authentic images are repurposed with misleading textual context. The dataset, drawn from four major news sources via the VisualNews corpus, has since become a standard evaluation benchmark for multimodal misinformation detection systems. Its inclusion as a supplementary evaluation dataset for MACE would strengthen the generalisability of findings beyond the Twitter-based benchmarks used in DANES and GETAE, and provide additional insight into the performance of the multimodal content branch when visual-textual consistency is deliberately compromised.

\subsection{Comprehensive Survey of Deep Learning Approaches}

Hu et al.\ \cite{Hu2022} provide a comprehensive survey of deep learning methods for fake news detection, systematically reviewing approaches spanning text-only content models, multimodal frameworks, social context architectures, and knowledge-enhanced systems. The survey identifies the integration of multimodal content with social propagation signals as an open research challenge, and highlights the absence of controlled ablation studies that isolate the contribution of each component as a key limitation of existing work. This observation directly motivates the comparative experimental design adopted by MACE, wherein controlled substitution of the content branch and ablation-based evaluation are employed to decompose the relative contributions of multimodal features and social context signals. The survey further confirms that the research questions addressed by this project — evaluating the impact of multimodal substitution within context-aware ensemble architectures — represent a genuine and unresolved gap in the current literature.

\subsection{Summary of Additional Literature Contributions}

The papers reviewed in this chapter complement the existing literature review by addressing dimensions not fully covered in the primary survey. Collectively, they reinforce three key arguments underlying the MACE framework: first, that temporal and propagation-based social context signals provide strong classification value independent of content modality \cite{Liu2020, Bian2020}; second, that cross-modal consistency and shared multimodal representations enhance fake news detection beyond what either modality achieves independently \cite{Zhou2020, Khattar2019}; and third, that the controlled evaluation of multimodal substitution within context-aware architectures represents a recognised and unaddressed gap in the literature \cite{Luo2021, Hu2022}. Together, these contributions strengthen the motivation for the MACE framework and provide additional benchmarking context for its experimental evaluation.

% ---------------------------------------------------------------
\section{Reflection on Evaluation Plan}
% ---------------------------------------------------------------

The evaluation plan outlined in Section 5 focuses on benchmarking MACE against DANES, GETAE, HMCAN, MPFN, DPSG, and Dellys et al.\ across FakeNewsNet and MuMin datasets. The additional literature reviewed in this chapter suggests several opportunities to further strengthen the evaluation design.

First, SAFE \cite{Zhou2020} and MVAE \cite{Khattar2019} represent established multimodal baselines that are well-cited and widely used in comparative evaluations across the fake news detection literature. Including them as additional benchmarks in the multimodal-without-social-context domain (alongside HMCAN and MPFN) would broaden the comparison set and provide a more complete picture of where MACE's content branch sits relative to prior multimodal work.

Second, Bi-GCN \cite{Bian2020} offers a strong propagation-based baseline that is directly evaluated on Twitter15 and Twitter16 — the same datasets used in MACE Stage 1. Its inclusion as an additional social-context baseline would allow for a more granular comparison of the DANES and GETAE social branches against an independent graph-based rumour detection model, strengthening the evidence base for retaining the propagation-graph social context branch in the full MACE framework.

Third, the NewsCLIPpings dataset \cite{Luo2021} presents a compelling supplementary evaluation set, as it specifically targets out-of-context multimodal misinformation — a category that is structurally distinct from the fabricated content in Twitter15/16 and FakeNewsNet. Evaluating MACE on NewsCLIPpings as an additional held-out benchmark would test the generalisability of the multimodal content branch to real-world misuse patterns where images are authentic but contextually misleading, a scenario not well represented by existing benchmark datasets.

Finally, the survey by Hu et al.\ \cite{Hu2022} notes that statistical significance testing and reporting of variance across runs remain inconsistently applied in the fake news detection literature. This reinforces the importance of the McNemar's test and paired t-test procedures already planned in the MACE evaluation framework, and suggests that reporting mean and standard deviation across multiple runs should be explicitly prioritised to ensure the reproducibility and credibility of results.

% ---------------------------------------------------------------
\section{Additional Literature Review Summary}
% ---------------------------------------------------------------

Table~\ref{tab:athul_lit_review} extends the Literature Review Summary Table (Table A.1) with the six additional studies reviewed in this chapter.

\begin{table}[ht]
\centering
\caption{Additional studies on multimodal and context-aware misinformation detection.}
\label{tab:athul_lit_review}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|p{3.5cm}|p{5.0cm}|p{4.5cm}|}
\hline
\textbf{Study and focus} & \textbf{Key finding / contribution} & \textbf{Main limitation} \\
\hline
Khattar et al.\ \cite{Khattar2019} -- MVAE \newline Multimodal variational autoencoder &
Demonstrated that generative multimodal modelling via a bimodal VAE outperforms discriminative baselines by up to 6\% accuracy on Twitter and Weibo &
Does not incorporate social context or propagation signals; evaluated only on content-level features \\
\hline
Liu \& Wu \cite{Liu2020} -- FNED \newline Fake news early detection &
Showed that temporal crowd-response patterns enable accurate detection within minutes of publication, achieving over 90\% accuracy on Twitter and Weibo &
Relies on early user response data which may be sparse for low-engagement posts; no visual modality \\
\hline
Bian et al.\ \cite{Bian2020} -- Bi-GCN \newline Bi-directional propagation graphs &
Established that encoding both top-down and bottom-up propagation structures significantly outperforms unidirectional graph models on Twitter15, Twitter16, and Weibo &
Requires full propagation tree structure; performance degrades in early detection scenarios with incomplete graphs \\
\hline
Zhou et al.\ \cite{Zhou2020} -- SAFE \newline Image-text similarity modelling &
Showed that explicit similarity scoring between text and image modalities improves fake news detection and provides interpretable cross-modal inconsistency signals &
Similarity-based approach may underperform when both modalities are individually coherent but jointly misleading \\
\hline
Luo et al.\ \cite{Luo2021} -- NewsCLIPpings \newline Out-of-context multimodal benchmark &
Introduced a large-scale benchmark targeting out-of-context image-caption mismatches, now a standard evaluation dataset for multimodal misinformation detection &
Dataset focuses on a specific misinformation subtype; does not include social context or propagation structure \\
\hline
Hu et al.\ \cite{Hu2022} \newline Deep learning survey &
Provided a comprehensive review identifying the integration of multimodal content and social context as an unresolved gap, directly motivating the MACE research questions &
As a survey, does not propose a new model or resolve the identified gap experimentally \\
\hline
\end{tabular}
\end{table}

% ---------------------------------------------------------------
\section*{Generative AI Usage}
\addcontentsline{toc}{section}{Generative AI Usage}
% ---------------------------------------------------------------

\subsection*{Declaration}

This contribution was primarily human-created. AI was used to support literature discovery, citation verification, and writing clarity. AI-generated content was reviewed, verified against primary sources, and approved before inclusion. The following model was used: Claude AI (Anthropic), Claude Sonnet.

\begin{table}[ht]
\centering
\caption{Generative AI usage report — Claude AI (Athul's contribution)}
\label{tab:athul_genai}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|p{4.5cm}|p{8.5cm}|}
\hline
\textbf{Field} & \textbf{Details} \\
\hline
1. GenAI tool used & Claude AI (Anthropic), Claude Sonnet 4.6 \\
\hline
2. Purpose for this assignment & Literature discovery and verification, drafting and refining written sections, checking citation accuracy, and identifying research papers relevant to the MACE framework not already included in the team's bibliography. \\
\hline
3. Prompts or instructions given & Representative prompts included: ``Find research papers on multimodal fake news detection not already in our bibliography''; ``Verify these 6 paper citations exist and are correctly attributed''; ``Write a LaTeX extended literature review section using these papers in the style of the existing proposal''; ``Suggest how these papers could strengthen the evaluation plan''. \\
\hline
4. Number of prompts given & Approximately 15--20 prompts across the literature discovery, verification, and writing stages. \\
\hline
5. Summary of GenAI output & Claude suggested candidate papers and their citations, verified them against academic databases, drafted LaTeX sections for the extended literature review and evaluation reflection, and formatted the bibliography entries in BibTeX. \\
\hline
6. Verification of GenAI output & All suggested paper citations were independently verified against Google Scholar, ACL Anthology, ACM Digital Library, and DBLP before inclusion. One suggested citation was found to be fabricated and replaced with a verified alternative. All DOIs were confirmed as valid. \\
\hline
7. Usage or modification of output & LaTeX drafts were reviewed and edited for accuracy, tone, and alignment with the existing proposal. All research arguments, connections to MACE, and analytical claims were evaluated and approved. No AI-generated content was included without human review. \\
\hline
8. Reflection & Useful for accelerating literature discovery and drafting, but required significant human oversight throughout. One hallucinated citation was identified and replaced through independent verification across multiple academic databases. AI-generated drafts were substantially reviewed and edited to ensure accuracy, contextual relevance, and alignment with the research proposal. The tool supported the process but did not replace critical thinking or scholarly judgement. \\
\hline
\end{tabular}
\end{table}
